### 2025-05-15 docs/article/Sections/state-of-the-art.tex
Chapter 2 - Distinguish template idioms based on DSL classification: external versus internal. Examples of external DSLs. Describe internal DSL and its advantages.

\textit{Web templates} have been the most widely adopted approach for
constructing dynamic HTML pages.
Web templates (such as JSP, Handlebars, or Thymeleaf), also known as a
\textit{web pages} or \textit{web views}~\cite{Fowler02,Alur01}, are based on
HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
\texttt{\{\{\}\}}, or \texttt{\$\{\}}), which represent \textit{dynamic}
information to be replaced at runtime with the results of corresponding
computations, producing the final HTML page.
The process of parsing and replacing these markers---i.e.,
\textit{resolution}---is the primary responsibility of the \textit{template
engine}~\cite{Parr04}.

@@

>>>>

\textit{Web templates} have been the most widely adopted approach for
constructing dynamic HTML pages.
Web templates (such as JSP, Handlebars, or Thymeleaf), also known as
\textit{web views}~\cite{Fowler02,Alur01}, are based on
HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
\texttt{\{\{\}\}}, or \texttt{\$\{\}}), which represent \textit{dynamic}
information to be replaced at runtime with the results of corresponding
computations, producing the final HTML page.
The process of parsing and replacing these markers---i.e.,
\textit{resolution}---is the primary responsibility of the \textit{template
engine}~\cite{Parr04}.

@@

Web templates are based on a \textit{domain-specific language}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
documents.
The DSL constrains the template's syntax and semantics to match the structure
and purpose of HTML.

DSLs can be divided in two types: \textit{external} or
\textit{internal}\cite{dslbook}.
\textit{External} DSLs are languages created without any affiliation to a
concrete programming language.
An example of an \textit{external} DSL is the regular expressions search
pattern\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. 
On the other hand an \textit{internal} DSL is defined within a host programming
language as a library and tends to be limited to the syntax of the host language,
such as Java.
JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
manipulation.

Traditionally, web template technologies use an \textit{external} DSL to define
control flow constructs and data binding primitives. Early web template engines
such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
approach as their templating dialect.
For example, a \texttt{foreach} loop can be expressed in each technology using its 
own DSL:
\\\texttt{<\% for(String item : items) \%>} in JSP, \texttt{<\% For Each item 
In items \%>} in legacy ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in 
Velocity, or \texttt{<?php foreach (\$items as \$item):?>} in PHP.

On the other hand, \textit{internal} DSLs for HTML allow templates to be defined 
directly within the \emph{host} language (such as Java, Kotlin, Groovy, Scala, or 
other general-purpose programming languages), rather than using text-based 
template files~\cite{carvalho2019hot}. In this case, a web template is not limited 
to templating constructs but may instead leverage any available primitive of the 
host language or any external API.

For example, in Java, you may iterate over data using a \texttt{for} statement,
or using the \texttt{java.util.streams} API, or any other JVM library such as
Guava, Varv, StreamEx, or others—according to your preferences.

Using an internal DSL can have several benefits over using textual templates:
\begin{enumerate}
  \item \emph{Type safety}: Because the templates are defined with the host
  programming language, the compiler can check the syntax and types of the
  templates at compile time, which can help catch errors earlier in the
  development process.

  \item \emph{IDE support}: Many modern IDEs provide code completion, syntax
  highlighting, and other features, which can make it easier to write and
  maintain templates.

  \item \emph{Flexibility}: Use all the features of the host programming language
  to generate HTML, can make it easier to write complex templates and reuse code.

  \item \emph{Integration}: Because the templates are defined in Java code, for
  example, you can easily integrate them with other Java code in your
  application, such as controllers, services, repositories and models.

\end{enumerate}
